Effects of the body force on the evacuation dynamics

### 2020-01-09 paper.tex
agregue mini conclusion en la parte de bottleneck

\begin{table}
\begin{tabular}{c@{\hspace{6mm}}c@{\hspace{6mm}}c@{\hspace{6mm}}c@{\hspace{6mm}}
c@{\hspace{14mm}}l}
 \hline
 $\tau$[s]   & $m$[kg]     & $v_d$[m/s]  &  $B$[m]  & $k$[kg/s$^2$] &  Refs. \\
 \hline
0.61         & ---         & 1.24 & $0.36+1.06\,v$ &  ---                 &  
 \cite{seyfried_2007} \\
0.50$^*$     & 80$^*$      & 1.34 & 0.50           &  ---                 &  
\cite{weidmann_1992,lakoba_2005}\\
---          & $67.5$      & 1.39 &  ---           &  $96.1 + 12694.1\,x$ & 
\cite{song_2019}\\
---          & $67.0$      & 1.39 &  ---           &  $97.0 + 29378.9\,x$ & 
\cite{song_2019}\\

@@

The reports from Refs.~\cite{seyfried_2007,weidmann_1992} do not include any 
values for the compressibility $k$ since these experiments were carried out 
under low density conditions. Notice that the minimum (perceived) step distance 
is $0.36\,$m \cite{seyfried_2007}, but the pedestrians seem to require larger 
distances when they walk faster. The commonly accepted value $B\simeq 
0.5\,$m is somewhat valid for walking speeds under $0.5\,$m/s 
\cite{seyfried_2007}. Higher walking speeds (say, $1\,$m/s) will require 
a step distance of $1.3\,$m for the pedestrians to feel that there is enough 
space to move along.     \\

The reported data from Ref.~\cite{song_2019} correspond to the crowded 
environment of the Beijing subway. This environment was not suitable for 
providing information on the step distance $B$, but estimates for the 
desired speed and the body compresibility could be achieved. The reported 
magnitude $k$ assesses either the clothes and the body compressibility. The 
final value, though, is linearly related to the compression $x$. \\

@@

The maximum (realistic) compressions may be computed from the relation 
$F=k(x)\,x$ and the compressibility $k(x)$ reported in Table~\ref{table_data}. 
An ``uncomfortable'' body force of $30\,$N can address compressions in the 
range of $0.030-0.055\,$m. Also, a ``hitting'' force of $60\,$N can address 
compressions between $0.045$ and $0.065\,$m. Thus, according to 
Table~\ref{table_data}, we may expect experimental values for $k(x)$ up to 
$1000\,$kg/s$^2$ (for $F=30\,$N), or, up to $1400\,$kg/s$^2$ (for $F=60\,$N). \\

@@

The sliding friction occurs in the tangential direction while the body force 
occurs in the normal direction. Both are assumed to be linear with respect to 
the net distance between contacting pedestrians. The sliding friction is also 
linearly related to the difference between the (tangential) velocities. The 
coefficients $\kappa$ (for the sliding friction) and $k$ (for the 
body force) are supposed to be related to the areas of contact and the clothes 
material, among others. \\

We stress that the expression (\ref{eqn_friction}) assumes fixed values for 
$\kappa$ and $k$. This may not be completely true according to 
Table~\ref{table_data}. The local density (and thus, the pedestrians' 
compression) may affect the compressibility parameter $k$ by more than an order 
of magnitude. We will vary $k$ (and $\kappa$) in order to explore this 
phenomenon. \\

@@

We confirm from these diagrams that no univocal relations can be established 
between the individual parameters and the collective dynamics (in a crowded 
environment). The presence of the body force moves the dynamics to a more 
complex context. We will investigate this context in Section~\ref{results}. \\

@@

We present in this Section the results corresponding to the bottleneck geometry. We show the consequences of modifying the body force coefficient $k$ in the evacuation dynamics. Recall that this coefficient is associated with the stiffness of the pedestrians. \\

Fig. \ref{vd_vs_te} shows the evacuation time as a function of the pedestrian's desired velocity. The evacuation time is defined as the time until 80\% of the pedestrians left the room. The different markers correspond to different $k$ values (\textit{i.e.} different body ``stiffness''; see caption for details). We will focus on the evacuation time for $v_d>2\,$m/s.\\

\begin{figure}[htbp!]
\centering
\includegraphics[width=0.7\columnwidth]
{./vd_vs_te_N225.eps}
\caption{\label{vd_vs_te}Mean evacuation time (s) vs. the pedestrian’s desired velocity (m/s) for a bottleneck. The room was 20 m x 20 m size. The door was 0.92 m width. Mean values were computed from 10 evacuation processes. 225 pedestrians were initially placed in a square lattice with a random initial velocity. Each process was finished when 158 pedestrians left the room. The different symbols indicate the $k$ value corresponding to the body force (see the label). The crosses correspond to the Helbing original SFM parameter, the up-triangles correspond to the value measured in Ref ..., squares correspond to zero body force and circles correspond to an extreme value of stiffness (one order of magnitude higher than the original SFM). The down triangles correspond to an intermediate value between the empirical value presented in Ref... and the chosen by Helbing in Ref.~\cite{helbing_2000} }
\end{figure}

Two behavioral patterns can be seen in Fig.~\ref{vd_vs_te}. Whenever the slope is positive it means that the harder the pedestrians try to get out (higher $v_d$), the higher it takes them to evacuate. This is the Faster-is-slower (FIS) regime. Conversely, the negative slope corresponds to a Faster-is-Faster (FIF) phenomenon (the harder they try, the quicker they leave). Fig.~\ref{vd_vs_te} shows a change in the slope sign only for stiffness values below $k=1.2$~E5. This mean that ``soft'' individuals are required to attain this behavioral change.   Notice that for  $k=1.2$~E6 (the stiffest explored pedestrians) no FIS can be seen at all. Besides, the evacuation pattern for $k = 0$ and $k = 1.2$~E4 are very similar since the body force is comparable to the social force for these stiffness values. \\

Despite the presence of the FIS or FIF phenomenon, the evacuation time at a fixed value of $v_d$ decreases for increasing values of $k$. This means that stiffer pedestrians evacuate faster than soft pedestrians. In order to investigate this issue, we computed the mean velocity of the whole crowd as a function of the stiffens $k$. Fig.~\ref{kn_vs_vx_bottleneck} shows the mean velocity in the x-direction as a function of the stiffness $k$ for three different desired velocities. In all cases, the velocity increases with the stiffness. From $k_n=0$ to $k_n=$1.2~E4 the increment of velocity is not as significative as it is for higher $k_n$. \\

\begin{figure}[!htbp]
\centering
    \subfloat[]{\includegraphics[width=0.7\columnwidth]{./kn_vs_vx_bottleneck.eps}\label{kn_vs_vx_bottleneck}}\\
\caption[width=0.47\columnwidth]{}
\end{figure}

The presence of a FIF phenomenon for only (soft individuals) open many questions on the microscopic dynamics of pedestrians. We may presume that contacts between pedestrians are quite different for soft individuals than for stiff individuals. In order to study the contacts dynamics, regardless of the compression effects, we assimilate the pedestrians as nodes and the whole crowd as a network. We set a link between any two individuals if both are in physical contact (\textit{i.e.} $r_{ij} \leq R{ij}$).\\

Fig.~\ref{degree_vd} shows the mean degree of the contact network as a function of the desired velocity. The degree of a node is defined as the number of links that connecting this node to any other node. This means, the number of pedestrians that are in physical contact with a given pedestrian. The mean degree is the average of the degree over all the nodes (pedestrians) and over the whole sampled time. We report mean values after the system has reached the stationary state, that is, after a well formed bulk has been established.\\

Increasing $v_d$ increases the mean degree, as expected. This occurs because increasing $v_d$ produces higher densities that force individuals to touch each other. But surprisingly. for a given $v_d$ the mean degree reduces as the $k$ value increases. A noticeable decrease of the mean degree can be seen for the highest explored value of $k$. We may expect a significant change in the sliding friction due to this fact.\\

A more detailed insight to the contact dynamics can be acquired from Fig.~\ref{overlap_vd}. The overlap between individuals is shown as a function of $v_d$. Recall from Section \ref{sfm} that the overlap is defined as $R_{ij}-r_{ij}$ where $R_{ij}$, is the sum of radius of particle $i$ and particle $j$ and $r_{ij}$  is the distance between both particles. Except for very low desired velocities (say, $v_d<2\,$m/s), we can see that the mean overlap is an increasing function of $v_d$ (for the studied $k$ values).\\

@@

\begin{figure}[!htbp]
\centering
    \subfloat[]{\includegraphics[width=0.45\columnwidth]{./degree_vs_vd_multi_kn.eps}\label{degree_vd}}\ 
    \subfloat[]{\includegraphics[width=0.45\columnwidth]{./overlap_vs_vd_multi_kn.eps}\label{overlap_vd}}\\
\caption[width=0.47\columnwidth]{(a) Mean degree as a function of the pedestrian’s desired velocity (m/s). (b) Mean overlap as a function of the pedestrians desired velocity. Each symbol indicates the $k$ value corresponding to the body force (see the labels). The data corresponds to a bottleneck with periodic boundary conditions (re-injecting pedestrians). The average was taken every two seconds once the crowd reaches the stationary state ($t=$20~s) until the end of the simulation ($t=$110~s).}
\label{degree_overlap_vd}
\end{figure}

@@

Besides, for a given $v_d$, the overlap increases as the $k$ value decreases. Reducing $k$ means reducing the normal body force which allows a greater invasion of the personal space of the individuals ( this means more overlap between pedestrians). The increment in mean overlap, however, seems to be more significant than the mean degree. \\

An increase in the body stiffness is expected to impact on the evacuation dynamics. The simulations reduction of the overlap and the mean degree between pedestrians, tend to diminish the sliding friction within the crowd. Consequently, an enhancement in the evacuation time (pedestrians escape faster) should be observed, as pictured in Fig.~\ref{vd_vs_te}. Notice, however, that switching from a FIS regime (positive slope) to a  FIF regime (negative slope) in Fig.~\ref{vd_vs_te} appears as a more complex phenomenon. We will focus on this issue in an upcoming investigation.\\

Despite the slidding friction, the body force has a notorious impact in the number of pedestrians touching each other (say, the degree). This is clearly depicted in Fig.~\ref{network_bottleneck} where four different configurations of the evacuation dynamics are shown. The configurations represent 225 pedestrians trying to escape through a door see caption for details. The colors correspond to the degree of each node (pedestrian) and the lines between pedestrians represent the contacts among them.\\

\begin{figure}[!htbp]
\centering
    \subfloat[]{\includegraphics[width=0.45\columnwidth]{./network_vd2_kn0.eps}\label{network_vd2_kn0}}\ 
    \subfloat[]{\includegraphics[width=0.45\columnwidth]{./network_vd10_kn0.eps}\label{network_vd10_kn0}}\\
        \subfloat[]{\includegraphics[width=0.45\columnwidth]{./network_vd2_kn1200000.eps}\label{network_vd2_kn1200000}}\ 
    \subfloat[]{\includegraphics[width=0.45\columnwidth]{./network_vd10_kn1200000.eps}\label{network_vd10_kn1200000}}\\
\caption[width=0.47\columnwidth]{Contact networks of a 225 pedestrian evacuation through a bottleneck. The door is placed at $(x,y)=(20,10)$~m, the width of the door is 0.92~m (equivalent to 2 pedestrian\textsc{\char13}s diameter). The lines that connect the nodes (pedestrians) represent the contact between them. The color represents the degree (the number of pedestrians with which it is connected). (a) and (b) correspond to a simulation without body force with $v_d=$2 and $v_d=$10 respectively. (c) and (d) correspond to simulations with $k$=1.2~E6 with $v_d=$2 and $v_d=$10 respectively.}
\label{network_bottleneck}
\end{figure}

The four configurations corresponds to two different $v_d$ and two different $k$ values (say, the  minimum and maximum explored values). Fig.~\ref{network_vd2_kn0} and Fig.~\ref{network_vd10_kn0} shows a snapshot for $k=$0, at desired velocities of 2$\,$m/s and 10$\,$m/s respectively.    Fig.~\ref{network_vd2_kn1200000} and Fig.~\ref{network_vd10_kn1200000} show similar situations, but for  $k=$1.2$\,$E6. As expected, increasing the desired velocity compresses the crowd towards the exit. \\

The four snapshots in \ref{network_bottleneck} comfirm (visually) the fact that more rigid bodies releases the crowd environment, widening the occupied region. At $v_d=10\,$m/s (the maximum explored velocity), it can hardly be found pedestians with degree 6 when $k=1.2\,$E6, while a lot of them are present for $k$=0.\\

\textcolor{red}{Agregar una conclusion}

@@

In this section, we present the results corresponding to the corridor geometry. We show the effects of modifying the body force coefficient $k$ in the dynamics.\\

@@

It is worth mentioning that increasing the body force reduces the flow, for a given density and a given friction value (compare \ref{flow_density_no_body_force} with \ref{flow_density_body_force} ). In other words, pedestrians walk slower along the corridor if the $k$ value increases (\textit{i.e.} if they are stiffer). This result is the opposite of what we get in bottleneck geometry since in bottleneck, the stiffer the pedestrians are, the faster they walk (and evacuate).\\

\begin{figure}[!htbp]
\centering
    \subfloat[]{\includegraphics[width=0.45\columnwidth]{./flow-density_multifric_nobodyforce.eps}\label{flow_density_no_body_force}}\ 
    \subfloat[]{\includegraphics[width=0.45\columnwidth]{./flow-density_multifric_bodyforce.eps}\label{flow_density_body_force}}\\
\caption[width=0.47\columnwidth]{Flow vs. global density. The flow is calculated in a circular area of R=1~m at the center of the corridor. The circular markers correspond to the original friction of the SFM, the "+" symbol corresponds to the friction increased by a factor of five and the triangles correspond to the friction increased by a factor of ten. The desired velocity was $v_d=1$ in all the cases. (a) corresponds to simulations without body force and (b) corresponds to a body force with $k=$1.2~E5.}
\label{flow_density}
\end{figure}

@@

Fig.~\ref{degree_dens} shows the mean degree as a function of the global density for different $k$ values. When the density is low,  the degree is zero because the pedestrians do not touch each other. When the density is around 4.5, a few pedestrians start to touch each other increasing the mean degree. As the density increases, the mean degree increases too until reaching an asymptotic behavior in degree six. When the degree is six, it corresponds to the densest packing possible for identical hard circles, the result is a hexagonal packing arrangement.\\

@@

The mean overlap increases as the global density increases (see Fig. \ref{overlap_dens}) and the higher the stiffness the lower the overlap. This result is similar to the overlap vs. desired velocity obtained for bottleneck geometry. The higher the value of $k$ the higher the repulsion of the pedestrian-pedestrian interaction since the body force is acting in the normal direction $n_{ij}$.\\

\begin{figure}[!htbp]
\centering
    \subfloat[]{\includegraphics[width=0.45\columnwidth]{./degree_vs_dens_multi_kn.eps}\label{degree_dens}}\ 
    \subfloat[]{\includegraphics[width=0.45\columnwidth]{./overlap_vs_dens_multi_kn.eps}\label{overlap_dens}}\\
\caption[width=0.47\columnwidth]{(a) Mean degree as a function of the global density for different $k$ values. (b) mean overlap as a function of the global density. The mean values are averages over all the pedestrians and over time once the system reaches the stationary state. Both measurements correspond to corridor geometry with desired velocity $v_d=1$.}
\label{network_corridor}
\end{figure}

As was mentioned above, increasing the $k$ value of the body force increases the stiffness of pedestrians. Increasing the stiffness reduces the overlap (both in bottleneck and corridor). Regarding the network connectivity, increasing the stiffness reduces the mean degree in a bottleneck but increases the mean degree in a corridor. In Fig.~\ref{network_corridor} it shown the contact network corresponding to the corridor at a particular timestep and a particular global density. Fig.~\ref{network_d6_kn0} corresponds to $k=0$ while Fig. \ref{network_d6_knE5} corresponds to $k=1.2$~E5. This is an example to show that increasing the stiffness increases the number of contacts between pedestrians (the network connectivity). Notice that there are many more pedestrians with degree 6 in the case of $k=1.2$~E5 than the case of $k=0$. \\

@@

\begin{figure}[!htbp]
\centering
    \subfloat[]{\includegraphics[width=0.45\columnwidth]{./network_d6_kn0.eps}\label{network_d6_kn0}}\ 
    \subfloat[]{\includegraphics[width=0.45\columnwidth]{./network_d6_knE5.eps}\label{network_d6_knE5}}\\
\caption[width=0.47\columnwidth]{Contact network of the pedestrians along the corridor. The lines that connect the nodes (pedestrians) represent the contact between them. The colors stand for the degree of the node (the number of pedestrians that are in contact with him/her). The corridor was 28~m $\times$ 22~m with periodic boundary conditions and $v_d=1$. (a) corresponds to a simulation without body force and (b) corresponds to a simulation with $k=1.2$~E5.}
\label{network_corridor}
\end{figure}

The increment of the connectedness is responsible for the reduction of the flow. In the corridor geometry, the crowd gets more connected if the pedestrians become stiffer. We studied the network properties of the contact graph arising from the positions $(x,y)$ of every pedestrian. We found that a good descriptor to exhibit the connectivity is the number of triangles per node. A triangle is defined as a structure composed of three nodes all in contact with each other. Fig.~\ref{triangles} shows the triangles per node as a function of the global density for different stiffness values $k$. This descriptor allows to distinguish states with the same global density but different stiffness levels. In other words, it is a descriptor very sensitive to changes in stiffness that allow expressing the connectivity of the system even better than the mean degree. \\

@@

\begin{figure}[htbp!]
\centering
\includegraphics[width=0.7\columnwidth]
{./triangles.eps}
\caption{\label{triangles} Triangles per node as a function of the global density. A triangle in a network is defined as three nodes all connected. Each marker corresponds to a different stiffness $k$ (see the label in the plot). The measurements correspond to a corridor of 28~m $\times$ 22~m with periodic boundary conditions and $v_d=1$.  }
\end{figure}

>>>>

\begin{table}
\begin{tabular}{c@{\hspace{6mm}}c@{\hspace{6mm}}c@{\hspace{6mm}}c@{\hspace{6mm}}
c@{\hspace{14mm}}l}
 \hline
 $\tau$[s]   & $m$[kg]     & $v_d$[m/s]  &  $B$[m]  & $k_n$[kg/s$^2$] &  Refs. \\
 \hline
0.61         & ---         & 1.24 & $0.36+1.06\,v$ &  ---                 &  
 \cite{seyfried_2007} \\
0.50$^*$     & 80$^*$      & 1.34 & 0.50           &  ---                 &  
\cite{weidmann_1992,lakoba_2005}\\
---          & $67.5$      & 1.39 &  ---           &  $96.1 + 12694.1\,x$ & 
\cite{song_2019}\\
---          & $67.0$      & 1.39 &  ---           &  $97.0 + 29378.9\,x$ & 
\cite{song_2019}\\

@@

The reports from Refs.~\cite{seyfried_2007,weidmann_1992} do not include any 
values for the compressibility $k_n$ since these experiments were carried out 
under low density conditions. Notice that the minimum (perceived) step distance 
is $0.36\,$m \cite{seyfried_2007}, but the pedestrians seem to require larger 
distances when they walk faster. The commonly accepted value $B\simeq 
0.5\,$m is somewhat valid for walking speeds under $0.5\,$m/s 
\cite{seyfried_2007}. Higher walking speeds (say, $1\,$m/s) will require 
a step distance of $1.3\,$m for the pedestrians to feel that there is enough 
space to move along.     \\

The reported data from Ref.~\cite{song_2019} correspond to the crowded 
environment of the Beijing subway. This environment was not suitable for 
providing information on the step distance $B$, but estimates for the 
desired speed and the body compresibility could be achieved. The reported 
magnitude $k_n$ assesses either the clothes and the body compressibility. The 
final value, though, is linearly related to the compression $x$. \\

@@

The maximum (realistic) compressions may be computed from the relation 
$F=k(x)\,x$ and the compressibility $k_n(x)$ reported in Table~\ref{table_data}. 
An ``uncomfortable'' body force of $30\,$N can address compressions in the 
range of $0.030-0.055\,$m. Also, a ``hitting'' force of $60\,$N can address 
compressions between $0.045$ and $0.065\,$m. Thus, according to 
Table~\ref{table_data}, we may expect experimental values for $k_n(x)$ up to 
$1000\,$kg/s$^2$ (for $F=30\,$N), or, up to $1400\,$kg/s$^2$ (for $F=60\,$N). \\

@@

The sliding friction occurs in the tangential direction while the body force 
occurs in the normal direction. Both are assumed to be linear with respect to 
the net distance between contacting pedestrians. The sliding friction is also 
linearly related to the difference between the (tangential) velocities. The 
coefficients $\kappa$ (for the sliding friction) and $k_n$ (for the 
body force) are supposed to be related to the areas of contact and the clothes 
material, among others. \\

We stress that the expression (\ref{eqn_friction}) assumes fixed values for 
$\kappa$ and $k_n$. This may not be completely true according to 
Table~\ref{table_data}. The local density (and thus, the pedestrians' 
compression) may affect the compressibility parameter $k_n$ by more than an order 
of magnitude. We will vary $k_n$ (and $\kappa$) in order to explore this 
phenomenon. \\

@@

We confirm from these diagrams that no univocal relations can be established 
between the individual parameters and the collective dynamics (in a crowded 
environment). The presence of the body force moves the dynamics to a more 
complex context. We will investigate this context in section~\ref{results}. \\

@@

We present in this section the results corresponding to the bottleneck geometry. We show the consequences of modifying the body force coefficient $k_n$ in the evacuation dynamics. Recall that this coefficient is associated with the stiffness of the pedestrians. \\

Fig.~\ref{vd_vs_te} shows the evacuation time as a function of the pedestrian's desired velocity. The evacuation time is defined as the time until 80\% of the pedestrians left the room. The different markers correspond to different $k_n$ values (\textit{i.e.} different body ``stiffness''; see caption for details). In this section we will focus on the evacuation time for $v_d>2\,$m/s.\\

\begin{figure}[htbp!]
\centering
\includegraphics[width=0.7\columnwidth]
{./vd_vs_te_N225.eps}
\caption{\label{vd_vs_te}Mean evacuation time (s) vs. the pedestrian’s desired velocity (m/s) for a bottleneck. The room was 20 m x 20 m size. The door was 0.92 m width. Mean values were computed from 10 evacuation processes. 225 pedestrians were initially placed in a square lattice with a random initial velocity. Each process was finished when 158 pedestrians left the room. The different symbols indicate the $k_n$ value corresponding to the body force (see the label). The crosses correspond to the Helbing original SFM parameter, the up-triangles correspond to the value measured in Ref.~\cite{gordon1989anthropometric}, squares correspond to zero body force and circles correspond to an extreme value of stiffness (one order of magnitude higher than the original SFM). The down triangles correspond to an intermediate value between the empirical value presented in Ref.~\cite{gordon1989anthropometric} and the chosen by Helbing in Ref.~\cite{helbing_2000} }
\end{figure}

Two behavioral patterns can be seen in Fig.~\ref{vd_vs_te}. Whenever the slope is positive, it means that the harder the pedestrians trying to get out (higher $v_d$), the higher it takes them to evacuate. This is the Faster-is-slower (FIS) regime. Conversely, the negative slope corresponds to a Faster-is-Faster (FIF) phenomenon (the harder they try, the quicker they leave). Fig.~\ref{vd_vs_te} shows a change in the slope sign only for stiffness values below $k=1.2$~E5. This means that ``soft'' individuals are required to attain this behavioral change.   Notice that for  $k_n=1.2$~E6 (the stiffest explored pedestrians), no FIS can be seen at all. Besides, the evacuation pattern for $k_n = 0$ and $k_n = 1.2$~E4 are very similar since the body force is comparable to the social force for these stiffness values. \\

Despite the presence of the FIS or FIF phenomenon, the evacuation time at a fixed value of $v_d$ decreases for increasing values of $k_n$. This means that stiffer pedestrians evacuate faster than soft pedestrians. To investigate this issue, we computed the mean velocity of the whole crowd as a function of the stiffens $k_n$. Fig.~\ref{kn_vs_vx_bottleneck} shows the mean velocity in the x-direction as a function of the stiffness $k_n$ for three different desired velocities. In all cases, the velocity increases with the stiffness. From $k_n=0$ to $k_n=$1.2~E4 the increment of velocity is not as significative as it is for higher $k_n$. \\

\begin{figure}[!htbp]
\centering
    \subfloat[]{\includegraphics[width=0.7\columnwidth]{./kn_vs_vx_bottleneck.eps}}\\
\caption[width=0.47\columnwidth]{Mean velocity in the x-direction as a function of the stiffness level $k_n$ for different desired velocities (see label). The data corresponds to a bottleneck with periodic boundary conditions (re-injecting pedestrians). The average was taken every two seconds once the crowd reaches the stationary state ($t=$20~s) until the end of the simulation ($t=$110~s). }
\label{kn_vs_vx_bottleneck}
\end{figure}

The presence of a FIF phenomenon only for soft individuals opens many questions on the microscopic dynamics of pedestrians. We may presume that contacts between pedestrians are quite different for soft individuals than for stiff individuals. To study the dynamics of the contact, regardless of the compression effects, we assimilate the pedestrians as nodes and the whole crowd as a network. We set a link between any two individuals if both are in physical contact (\textit{i.e.} $r_{ij} \leq R{ij}$).\\

Fig.~\ref{degree_vd} shows the mean degree of the contact network as a function of the desired velocity. The degree of a node is defined as the number of links that connect this node to any other node. This means, the number of pedestrians that are in physical contact with a given pedestrian. The mean degree is the average of the degree over all the nodes (pedestrians) and over the whole sampled time. We report mean values after the system has reached the stationary state, that is after a well-formed bulk has been established.\\

Increasing $v_d$ increases the mean degree, as expected. This occurs because increasing $v_d$ produces higher densities that force individuals to touch each other. But surprisingly, for a given $v_d$, the mean degree reduces as the $k_n$ value increases. A noticeable decrease in the mean degree can be seen for the highest explored value of $k_n$. We may expect a significant change in the sliding friction due to this fact.\\

A more detailed insight into the contact dynamics can be acquired from Fig.~\ref{overlap_vd}. The overlap between individuals is shown as a function of $v_d$. Recall from section \ref{sfm} that the overlap is defined as $R_{ij}-r_{ij}$ where $R_{ij}$, is the sum of radius of particle $i$ and particle $j$ and $r_{ij}$  is the distance between both particles. Except for very low desired velocities (say, $v_d<2\,$m/s), we can see that the mean overlap is an increasing function of $v_d$ (for the studied $k_n$ values).\\

@@

\begin{figure}[!htbp]
\centering
    \subfloat[]{\includegraphics[width=0.45\columnwidth]{./degree_vs_vd_multi_kn.eps}\label{degree_vd}}\ 
    \subfloat[]{\includegraphics[width=0.45\columnwidth]{./overlap_vs_vd_multi_kn.eps}\label{overlap_vd}}\\
\caption[width=0.47\columnwidth]{(a) Mean degree as a function of the pedestrian’s desired velocity (m/s). (b) Mean overlap as a function of the pedestrians desired velocity. Each symbol indicates the $k_n$ value corresponding to the body force (see the labels). The data corresponds to a bottleneck with periodic boundary conditions (re-injecting pedestrians). The average was taken every two seconds once the crowd reaches the stationary state ($t=$20~s) until the end of the simulation ($t=$110~s).}
\label{degree_overlap_vd}
\end{figure}

@@

Besides, for a given $v_d$, the overlap increases as the $k_n$ value decreases. Reducing $k_n$ means reducing the normal body force, which allows a more significant invasion of the personal space of the individuals ( this means more overlap between pedestrians). The increment in mean overlap, however, seems to be more significant than the mean degree. \\

Increasing body stiffness has a significant impact on evacuation dynamics. The simulations reduction of the overlap and the mean degree between pedestrians, tend to diminish the sliding friction within the crowd. Consequently, an enhancement in the evacuation time (pedestrians escape faster) is observed, as pictured in Fig.~\ref{vd_vs_te}. Notice, however, that switching from a FIS regime (positive slope) to a  FIF regime (negative slope) in Fig.~\ref{vd_vs_te} appears as a more complex phenomenon. We will focus on this issue in an upcoming investigation.\\

Despite the sliding friction, the body force has a notorious impact in the number of pedestrians touching each other (say, the degree). This is clearly depicted in Fig.~\ref{network_bottleneck} where four different configurations of the evacuation dynamics are shown. The configurations represent 225 pedestrians trying to escape through a door see the caption for details. The colors correspond to the degree of each node (pedestrian), and the lines between pedestrians represent the contacts among them.\\

\begin{figure}[!htbp]
\centering
    \subfloat[]{\includegraphics[width=0.45\columnwidth]{./network_vd2_kn0.eps}\label{network_vd2_kn0}}\ 
    \subfloat[]{\includegraphics[width=0.45\columnwidth]{./network_vd10_kn0.eps}\label{network_vd10_kn0}}\\
        \subfloat[]{\includegraphics[width=0.45\columnwidth]{./network_vd2_kn1200000.eps}\label{network_vd2_kn1200000}}\ 
    \subfloat[]{\includegraphics[width=0.45\columnwidth]{./network_vd10_kn1200000.eps}\label{network_vd10_kn1200000}}\\
\caption[width=0.47\columnwidth]{Contact networks of a 225 pedestrian evacuation through a bottleneck. The door is placed at $(x,y)=(20,10)$~m, the width of the door is 0.92~m (equivalent to 2 pedestrian\textsc{\char13}s diameter). The lines that connect the nodes (pedestrians) represent the contact between them. The color represents the degree (the number of pedestrians with which it is connected). (a) and (b) correspond to a simulation without body force with $v_d=$2 and $v_d=$10 respectively. (c) and (d) correspond to simulations with $k_n$=1.2~E6 with $v_d=$2 and $v_d=$10 respectively.}
\label{network_bottleneck}
\end{figure}

The four configurations corresponds to two different $v_d$ and two different $k_n$ values (say, the  minimum and maximum explored values). Fig.~\ref{network_vd2_kn0} and Fig.~\ref{network_vd10_kn0} shows a snapshot for $k_n=$0, at desired velocities of 2$\,$m/s and 10$\,$m/s respectively.    Fig.~\ref{network_vd2_kn1200000} and Fig.~\ref{network_vd10_kn1200000} show similar situations, but for  $k_n=$1.2$\,$E6. As expected, increasing the desired velocity compresses the crowd towards the exit. \\

The four snapshots in Fig.~\ref{network_bottleneck} confirm (visually) the fact that more rigid pedestrians release the crowded environment, widening the occupied region. At $v_d=10\,$m/s (the maximum explored velocity), it can hardly be found pedestrians with degree 6 when $k_n=1.2\,$E6, while a lot of them are present for $k_n$=0.\\

To conclude this section, we want to emphasize that increasing body stiffness has a significant impact on the evacuation dynamics. The stiffer pedestrians are, the less they touch to each other (less degree and less overlap). This yields a reduction of an increment of the speed, which reduces the evacuation time. \\

@@

In this section, we present the results corresponding to the corridor geometry. We show the effects of modifying the body force coefficient $k_n$ in the dynamics.\\

@@

It is worth mentioning that increasing the body force reduces the flow, for a given density and a given friction value (compare \ref{flow_density_no_body_force} with \ref{flow_density_body_force} ). In other words, pedestrians walk slower along the corridor if the $k_n$ value increases (\textit{i.e.} if they are stiffer). This result is the opposite of what we get in bottleneck geometry since in bottleneck, the stiffer the pedestrians are, the faster they walk (and evacuate).\\

\begin{figure}[!htbp]
\centering
    \subfloat[]{\includegraphics[width=0.45\columnwidth]{./flow-density_multifric_nobodyforce.eps}\label{flow_density_no_body_force}}\ 
    \subfloat[]{\includegraphics[width=0.45\columnwidth]{./flow-density_multifric_bodyforce.eps}\label{flow_density_body_force}}\\
\caption[width=0.47\columnwidth]{Flow vs. global density. The flow is calculated in a circular area of R=1~m at the center of the corridor. The circular markers correspond to the original friction of the SFM, the "+" symbol corresponds to the friction increased by a factor of five and the triangles correspond to the friction increased by a factor of ten. The desired velocity was $v_d=1$ in all the cases. (a) corresponds to simulations without body force and (b) corresponds to a body force with $k_n=$1.2~E5.}
\label{flow_density}
\end{figure}

@@

Fig.~\ref{degree_dens} shows the mean degree as a function of the global density for different $k_n$ values. When the density is low,  the degree is zero because the pedestrians do not touch each other. When the density is around 4.5, a few pedestrians start to touch each other increasing the mean degree. As the density increases, the mean degree increases too until reaching an asymptotic behavior in degree six. When the degree is six, it corresponds to the densest packing possible for identical hard circles, the result is a hexagonal packing arrangement.\\

@@

The mean overlap increases as the global density increases (see Fig. \ref{overlap_dens}) and the higher the stiffness the lower the overlap. This result is similar to the overlap vs. desired velocity obtained for bottleneck geometry. The higher the value of $k_n$ the higher the repulsion of the pedestrian-pedestrian interaction since the body force is acting in the normal direction $n_{ij}$.\\

\begin{figure}[!htbp]
\centering
    \subfloat[]{\includegraphics[width=0.45\columnwidth]{./degree_vs_dens_multi_kn.eps}\label{degree_dens}}\ 
    \subfloat[]{\includegraphics[width=0.45\columnwidth]{./overlap_vs_dens_multi_kn.eps}\label{overlap_dens}}\\
\caption[width=0.47\columnwidth]{(a) Mean degree as a function of the global density for different $k_n$ values. (b) mean overlap as a function of the global density. The mean values are averages over all the pedestrians and over time once the system reaches the stationary state. Both measurements correspond to corridor geometry with desired velocity $v_d=1$.}
\label{network_corridor}
\end{figure}

As was mentioned above, increasing the $k_n$ value of the body force increases the stiffness of pedestrians. Increasing the stiffness reduces the overlap (both in bottleneck and corridor). Regarding the network connectivity, increasing the stiffness reduces the mean degree in a bottleneck but increases the mean degree in a corridor. In Fig.~\ref{network_corridor} it shown the contact network corresponding to the corridor at a particular timestep and a particular global density. Fig.~\ref{network_d6_kn0} corresponds to $k_n=0$ while Fig. \ref{network_d6_knE5} corresponds to $k_n=1.2$~E5. This is an example to show that increasing the stiffness increases the number of contacts between pedestrians (the network connectivity). Notice that there are many more pedestrians with degree 6 in the case of $k_n=1.2$~E5 than the case of $k_n=0$. \\

@@

\begin{figure}[!htbp]
\centering
    \subfloat[]{\includegraphics[width=0.45\columnwidth]{./network_d6_kn0.eps}\label{network_d6_kn0}}\ 
    \subfloat[]{\includegraphics[width=0.45\columnwidth]{./network_d6_knE5.eps}\label{network_d6_knE5}}\\
\caption[width=0.47\columnwidth]{Contact network of the pedestrians along the corridor. The lines that connect the nodes (pedestrians) represent the contact between them. The colors stand for the degree of the node (the number of pedestrians that are in contact with him/her). The corridor was 28~m $\times$ 22~m with periodic boundary conditions and $v_d=1$. (a) corresponds to a simulation without body force and (b) corresponds to a simulation with $k_n=1.2$~E5.}
\label{network_corridor}
\end{figure}

The increment of the connectedness is responsible for the reduction of the flow. In the corridor geometry, the crowd gets more connected if the pedestrians become stiffer. We studied the network properties of the contact graph arising from the positions $(x,y)$ of every pedestrian. We found that a good descriptor to exhibit the connectivity is the number of triangles per node. A triangle is defined as a structure composed of three nodes all in contact with each other. Fig.~\ref{triangles} shows the triangles per node as a function of the global density for different stiffness values $k_n$. This descriptor allows to distinguish states with the same global density but different stiffness levels. In other words, it is a descriptor very sensitive to changes in stiffness that allow expressing the connectivity of the system even better than the mean degree. \\

@@

\begin{figure}[htbp!]
\centering
\includegraphics[width=0.7\columnwidth]
{./triangles.eps}
\caption{\label{triangles} Triangles per node as a function of the global density. A triangle in a network is defined as three nodes all connected. Each marker corresponds to a different stiffness $k_n$ (see the label in the plot). The measurements correspond to a corridor of 28~m $\times$ 22~m with periodic boundary conditions and $v_d=1$.  }
\end{figure}
